Legion: Expressing Locality and Independence with Logical Regions

### 2012-04-26 doc/sc2012/legion.tex
Working on the code example for the paper

\bibliographystyle{plainnat}
\usepackage{amsmath,amsthm}
\usepackage{listings}
\usepackage{tikz}
\usepackage{subfigure}
\usepackage{multirow}
\usepackage{pgfplots}
\usepackage{booktabs}
\usepackage{dcolumn}
\usepackage{cancel}

@@

\title{Legion: Expressing Locality and Independence with Logical Regions}
\author{Author names omitted}

@@

\begin{abstract}
%As transistor counts have continued to scale with Moore's Law, computer
%architectures are becoming increasingly parallel to avoid fundamental
%physical limitations.  
Modern parallel architectures have both
heterogeneous processors and deep, complex memory hierarchies.  
%New
%programming abstractions are required to grant programmers the power
%to leverage these architectures.  With this problem in mind 
We present Legion, 
a programming model, type system, and runtime for programming these machines.
Legion is organized around {\em logical regions}, which express both locality and independence of program data.  
Legion also enables explicit, programmer controlled movement of data through
the memory hierarchy and placement of tasks based on locality information
via a novel mapping interface.  Running on a 4 node cluster with 8 total GPUs 
and 4 levels of memory hierarchy, our implementation of Legion 
achieves up to 5.9X speedup over a single CPU-GPU node on real-world applications.
\end{abstract}

\section{Introduction}
\label{sect:intro}
Put intro text here \cite{Fatahalian06}

%\input{circuit}
%\input{code_ex}
%\input{part_fig}

>>>>

\bibliographystyle{IEEEtran}
\usepackage{amsmath,amsthm}
\usepackage{listings}
\usepackage{tikz}
\usepackage{subfigure}
\usepackage{multirow}
\usepackage{pgfplots}
\usepackage{booktabs}
\usepackage{dcolumn}
\usepackage{cancel}

@@

\pdfpagewidth=8.5in
\pdfpageheight=11in

\title{Legion: Expressing Locality and Independence with Logical Regions}
\author{
\IEEEauthorblockN{Michael Bauer\IEEEauthorrefmark{1},
%\authorinfo(Stanford University, email: \mbox{mebauer@cs.stanford.edu}
Sean Treichler\IEEEauthorrefmark{1}, 
%\authorinfo(Stanford University, email: \mbox{sjt@cs.stanford.edu}
Elliott Slaughter\IEEEauthorrefmark{1}, 
% \authorinfo(Stanford University, email: \mbox{slaughter@cs.stanford.edu}
Alex Aiken\IEEEauthorrefmark{1}  
%\authorinfo(Stanford University, email: \mbox{aiken@cs.stanford.edu}
%\email{aiken@cs.stanford.edu}
}
\IEEEauthorblockA{
\IEEEauthorrefmark{1}{Computer Science Department, Stanford University}
}
}

@@

\begin{abstract}
Modern parallel architectures have both heterogeneous processors and
deep, complex memory hierarchies.  We present Legion, a programming
model and runtime system for achieving high performance on these
machines.  Legion is organized around {\em logical regions}, which
express both locality and independence of program data and {\em
  tasks}, functions that perform computations on regions.  We describe
a runtime system that dynamically extracts parallelism from
Legion programs, using a distributed, parallel scheduling algorithm
that identifies both independent tasks and nested parallelism.  Legion
also enables explicit, programmer controlled movement of data through
the memory hierarchy and placement of tasks based on locality
information via a novel mapping interface.  We evaluate our Legion
implementation using four applications: fluid-flow on a regular grid,
a three-level AMR code solving a simple heat transfer equation, a circuit simulation,
and LINPACK.
%Running on a 4 node cluster with 8 total GPUs 
%and 4 levels of memory hierarchy, our implementation of Legion 
%achieves up to 5.9X speedup over a single CPU-GPU node on real-world applications.
\end{abstract}

\section{Introduction}
\label{sect:intro}
%Put intro text here \cite{Fatahalian06}

Modern parallel machines are increasingly complex, with deep,
distributed memory hierarchies and heterogeneous processing units.  On
these architectures, it is crucial for performance that the programmer
and the compiler be able to reason about {\em locality} (data resident
close to computation that uses it) and {\em independence} (computations
operating on disjoint data).  Most contemporary
languages have no facilities for the programmer to express locality
and independence.  The few languages that do focus primarily on
array-based locality \cite{Fatahalian06,CHAPEL04,UPC99} and 
avoid irregular pointer data structures.
%encode such structures in arrays, which 
%data structures encoded in arrays which are impossible for compilers
%and runtime systems to reason about with regard to locality and independence.

In this paper we describe Legion, a parallel programming system based
on using {\em logical regions} to describe the organization of data.
A logical region names a set of objects.  Logical regions are first
class values in Legion, and in particular may be passed as arguments
to functions that access the data in those regions, providing locality
information.  Logical regions may be {\em partitioned} into (usually)
disjoint {\em subregions}, providing information for determining independence of computations.  Furthermore,
computations access logical regions with particular {\em privileges} (e.g., {\em
  read-only} and {\em read-write}, among others) 
and {\em coherence} (e.g., {\em exclusive access} and {\em atomic access}, among others).

We begin with an extended example of a Legion program in Section~\ref{sec:ex}.
Each subsequent section describes a level in the Legion programming
system from the application down to the machine:

\begin{itemize}

\item We summarize Legion's static type system (Section~\ref{sec:type}), which serves two primary
  purposes.  First, pointers to regions are checked
  statically, eliminating the need for expensive runtime checks.
  Second, the invariant that a function $f$ accesses only
  regions passed as arguments to $f$, or subregions of those regions, is enforced.
  This property guarantees the Legion implementation can rely on
  preservation of independence---if two functions are passed disjoint
  region arguments, their computations access disjoint
data.

\item We describe the execution semantics for Legion and show that
analyzing region dependencies within a single function is sufficient to 
guarantee that a Legion execution preserves sequential program
behavior if regions are accessed with the strongest (exclusive)
coherence privilege (Section~\ref{sec:exec}).  This local scheduling property is central to 
the scalability of Legion on distributed memory machines.

\item We describe the design and implementation of the Legion {\em high-level runtime}
system (Section~\ref{sec:highlevel}), which performs scheduling of {\em tasks} (functions to be executed in parallel)
and implements each logical region as one or more {\em physical regions}.
The high-level runtime's scheduling algorithm is analogous to an out-of-order processor,
but works at the granularity of the tasks and regions in a function body rather than the instructions and registers
in a code block.  We describe the stages of the task execution pipeline, and how the high-level runtime coordinates parallel
scheduling decisions across the machine.  We also present the {\em mapping} interface, an API for incorporating
application-specific knowledge into the high-level runtime's decisions.

\item We describe the design and implementation of the Legion {\em low-level runtime} (Section~\ref{sec:lowlevel}),
a portability layer designed to abstract a wide variety of hardware, including multicore chips, clusters, and accelerators
such as GPUs.

\item We present the results of experiments on two applications running on a cluster of multicore processors with
attached GPUs (Section~\ref{sec:exp}).  The applications are both irregular and illustrate the capability of Legion to 
exploit locality and independence on such platforms.
Running on a 4 node cluster with 8 total GPUs 
and 4 levels of memory hierarchy, our implementation of Legion 
achieves up to 5.9X speedup over a single CPU-GPU node on these applications.

\end{itemize}
%\input{circuit}
\input{code_ex}
%\input{part_fig}
